---
title: "Analysis Of Damage Caused By Severe Weather Events"
author: "M Williams"
date: "16 August 2014"
output: html_document
---
##Synopsis  
Analysis of the damage caused by storms in the United States during the years 1950 to 2011. The analysis is done both in terms of material damage and caualties caused. The material damage consists of damage to property and to crops. Casualties are both fatalities and injuries. In both cases the top ten in terms of total are reported, along with a breakdown of their constituents.  

##Data Processing
Load required libraries  

```{r load-library, message=FALSE}
library(sqldf)
library(ggplot2)
library(tcltk)
```

Load data into a data frame  

```{r load-data, cache=TRUE}
storm <- read.csv(bzfile('StormData.csv.bz2'))
```  
Use sqldf to create summaries of the data by event type.  

```{r by-type, cache=TRUE}
byType <- sqldf('select EVTYPE, sum(FATALITIES) fatalities, sum(INJURIES) injuries, sum(FATALITIES+INJURIES) casualties,
                sum(PROPDMG) property, sum(CROPDMG) crops, sum(PROPDMG+CROPDMG) damage
                from storm
                group by EVTYPE order by EVTYPE')
```  
Select top ten human casualty data and create a data frame for the plot  

```{r}
cas <- sqldf('select EVTYPE, fatalities, injuries, casualties
                from byType
                where casualties > 0
                order by casualties desc')

topType <- cas[1:10,]

casPlt <- sqldf('select EVTYPE evtype, casualties, "Total" rtype
      from topType')

casPlt <- rbind(casPlt, sqldf('select EVTYPE evtype, fatalities casualties, "Fatality" rtype
      from topType'))

casPlt <- rbind(casPlt, sqldf('select EVTYPE evtype, injuries casualties, "Injury" rtype
      from topType'))

```  
Select top ten property damage data and create a data frame for the plot  

```{r}
eco <- sqldf('select EVTYPE, property, crops, damage
                from byType
                where damage > 0
             order by damage desc')

topTypeEco <- eco[1:10,]

ecoPlt <- sqldf('select EVTYPE evtype, damage, "Total" rtype
      from topTypeEco')

ecoPlt <- rbind(ecoPlt, sqldf('select EVTYPE evtype, property damage, "Property Damage" rtype
      from topTypeEco'))

ecoPlt <- rbind(ecoPlt, sqldf('select EVTYPE evtype, crops damage, "Crop Damage" rtype
      from topTypeEco'))
```
##Results  
###Human Damage
The events which are most harmful to population health are:  

```{r}
topType

ggplot(data=casPlt, aes(x=evtype, y=casualties, fill=factor(rtype))) + 
    ggtitle('Human Cost Of Storm Damage') +
    geom_bar(position="dodge", stat="identity") + coord_flip() +
    scale_x_discrete(limits=rev(topType$EVTYPE), name="Event") +
    scale_y_continuous(name="Casualties") + 
    theme(legend.title=element_blank(), legend.position=c(.75,.7))

```  
  
###Economic Damage
The events which cause the most economic damage are:  

```{r}
topTypeEco

ggplot(data=ecoPlt, aes(x=evtype, y=damage, fill=factor(rtype))) + 
    ggtitle('Economic Cost Of Storm Damage') +
    geom_bar(position="dodge", stat="identity") + coord_flip() +
    scale_x_discrete(limits=rev(topTypeEco$EVTYPE), name="Event") +
    scale_y_continuous(name="Cost") +
    theme(legend.title=element_blank(), legend.position=c(.75,.7))
```
